---
title: Spatio-temporal drivers of NPP 
author: Alexey Shiklomanov
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = TRUE, echo = FALSE, dev = "png")
```

# Introduction

Currently, Hector effectively has a single global plant functional type, with a single common set of parameters ($\beta$, the CO2 fertilization response, and the respiration-temperature Q10).
In reality, different ecosystems respond differently to changes in climate (including CO2, temperature, and precipitation).
The overarching objective is to find the minimum number of such plant functional types that captures some target fraction of the variance in climate-NPP response, and to potentially identify which factors besides CO2 explain the most variability in productivity trends across different biomes.

Traditionally, we select PFTs based on _a priori_ knowledge about differences between the structure and physiology of different biomes.
The idea of this analysis is to instead to an entirely data-driven approach -- i.e. to see if any obvious categories emerge out of analyses of spatio-temporal NPP data.
This document presents an initial exploration of such a dataset to provide a sense of how I'm thinking about approaching this problem.

# Data

So far, this analysis only uses one NPP dataset: the [Global Production Efficiency Model (GloPEM)](http://glcf.umd.edu/data/glopem/), developed by the University of Maryland.
It is based on optical observations of the NOAA's [Advanced Very-High Resolution Radiometer (AVHRR)](https://noaasis.noaa.gov/NOAASIS/ml/avhrr.html).
It has global coverage at 8 km spatial resolution, and is available in 10 day periods or annual sums (used here) from 1981 to 2000.

To get a sense of how the GloPEM data look, here is a map of the data for 2000, the last year they are available.

```{r plot_2000, fig.cap = "GloPEM total annual NPP in 2000.", cache = TRUE}
readd(glcf_2000_plot)
```

Here's the same map zoomed in on the United States.

```{r plot_2000_usa, fig.cap = "GloPEM total annual NPP in 2000.", warning = FALSE, cache = TRUE}
readd(glcf_2000_plot_usa)
```

For the ENSO analysis, I am using the monthly [NINO3 index](https://www.esrl.noaa.gov/psd/gcos_wgsp/Timeseries/Nino3/), averaged annually, calculated from the HadlSST1 reanalysis.
The dataset spans 1870 to present, but I am only using values that overlap the GloPEM data.
A time series of the index is shown below.

```{r plot_nino3, fig.cap = "Annually averaged NINO3 index"}
readd(nino3_plot)
```

# Empirical orthogonal function analysis

One way to reduce the dimensionality of this dataset is via "Empirical Orthogonal Functions", which are often used in climate science to identify patterns of spatio-temporal variability.
The analysis can be summarized in the following steps:

1. Flatten the 3D dataset (latitude x longitude x time) down to two dimensions (space x time).

2. Perform a principal component analysis (Eigen / singular value decomposition) on the resulting matrix.

3. Re-project the resulting PCA scores back onto the 2D coordinate space.

The GloPEM dataset contains 20 years of data.
However, because we are mostly interested in drivers of change in NPP, I remove the first year's NPP from every subsequent year, such that the analysis is on the change in NPP for every year starting in 1981.
This gives us 19 components to analyze.

The first step is to identify what fraction of variance is explained by each component.
Below is a plot of the cumulative variance fraction explained by the 20 components.

```{r plot_pca_sd, fig.cap = "Cumulative variance fraction explained by first _n_ principal components."}
readd(pca_sd_plot)
```

Unfortunately, this curve is fairly shallow, meaning that we need a lot of PCA components to explain the variance in NPP.
The first 10 components explain just under 70% of the variance.
Let's examine the time series associated with each component.

```{r plot_pca_ts, fig.cap = "Loadings of first 10 PCA components"}
readd(loadings_ts)
```

The first principal component is dominated by a monotonic increase (we know this is an increase _a priori_; the absolute sign of PC loadings is actually meaningless) in NPP.
This suggests that the simple `NPP = f(CO2)` formulation in Hector is actually capturing the most important component of the global NPP temporal trajectory.
However, based on the previous plot, we know that this first component only explains around 12.5% of the global NPP variance.
What are some factors that can explain the remaining components?

One way to tackle this question is to look at maps of the component scores.
For instance, below is a map of the scores for PC1.

```{r plot_pc1, fig.cap = "Projection of PC1 scores", cache = TRUE}
scores <- readd(scores)#[Comp.1 > 3000, Comp.1 := 3000][Comp.1 < -3000, Comp.1 := -3000]
plt <- ggplot(scores) +
  aes(x = lon, y = lat, fill = Comp.1) +
  geom_tile() +
  scale_fill_gradient2(low = "blue4", mid = "white", high = "red4")
plt
```

This map looks mostly white because the distribution of scores is really uneven.
For instance, there is a patch of extremely strong negative scores around the Nile River.

```{r plot_pc1_nile, fig.cap = "Projection of PC1 scores: Zoom on Nile River", cache = TRUE}
plt + xlim(27, 40) + ylim(20, 33)
```

Most of the scores are clustered between around -1500 and 1500.

```{r hist_pc1, fig.cap = "Histogram of PC1 scores"}
hist(scores[, Comp.1], main = "", xlab = "PC1 score")
```

Therefore, let's redraw the same figure, cropping the colors to the range `[-1500, 1500]`.

```{r plot_pc1_cropped}
plt %+% scores[abs(Comp.1) > 1500, Comp.1 := sign(Comp.1) * 1500]
```

This map is still very patchy, but it is a little easier to see some cohesive spatial trends.
For instance, NPP in the savannas at the southern edge of the Sahara desert show a strong decline in NPP, which are likely related to the desertification of that region.
Meanwhile, in the north-eastern and central Amazon, along the Atlantic coast of sub-equatorial Africa, and in central Asia, NPP is increasing over this period.

Because component scores are arbitrary, it's hard to understand exactly what's going on with the remaining components without looking at some ancillary data.
A known driver of variability in NPP is the El Niño Southern Oscillation (ENSO).
Using these components, we can ask: Which (if any) of these components are related to ENSO?
We can answer this by looking at temporal correlations with an ENSO index (see _Methods_).

```{r enso_correlation, fig.cap = "Temporal correlation of all PCA components with NINO3 index"}
readd(enso_corr) %>%
  ggplot() +
    aes(x = component, y = R) +
    geom_col() +
    labs(x = "PCA component", y = "Correlation with NINO3 index")
```

PC4 shows a strong positive correlation with ENSO.
What does its projection look like?

```{r enso_projection, fig.cap = "Projection of PC4"}
plt %+% aes(fill = Comp.4)
```
